BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2024-07-10 manuscript/manuscript_APPS.rmd
more citations

In fact innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative). 
The current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative). 
Further we anticipate that collections are yet to gain their most widespread utilization as a revitalization of natural history appears underway in ecology, fostered via novel approaches such as remote and electronic sensing, meta-barcoding, and community science  (@tosa2021rapid). 
While image or purely observational (rather than collections based) citizen science initializes (e.g. iNaturalist) have dovetailed to meet many needs of these studies specimens contain rich data which are not accessible via images. 
Namely specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes (@borges2020schrodinger), and seeds or pollen; eternally ensuing specimens as the ultimate data source to center most efforts around.

However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow. 
We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, than simply taking photographs via well-developed apps (@daru2018widespread, @mishler2020spatial). 
While many young botanists, capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria. 
Apparent problems include the lack of dedicated time at a field seasons end to process specimens, and a general lack of education on cartography and orienteering, natural history (e.g. geology, geomorphology), nomenclature and Latin, various computer programs (e.g. Microsoft Office suite), and increasingly - plant systematics. 
In the absence of suitable mentors this assuredly results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for many specimens to be accessioned at all, and increasingly ever collected.

@@

The generation of an herbarium specimen includes many steps which are easy to take for granted. 
For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate resources (printed topographic maps, or GIS software) at their disposal. 
In topographically complex areas, where borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their position. 
Finding appropriate sites names is another problem which can rarely be solved without a printed map, as many software maps now consider many features which would serve as site names extraneous in the era of GPS. 
The rate at which taxonomic innovations are occurring has left many Floras difficult for young users to interpret and has made it difficult for them to find more recently applied names.
Upon finding a name they may find it frustrating to hear that while published, the proposal has been accepted by few practitioners and they have unwittingly offended certain curators. 
Formatting a label correctly (e.g. abbreviations), if successful upon even setting up a mail merge, is a time consuming process and likely to introduce several errors in formatting. 
Even if a collector navigates all of these hurdles successfully, the time allocated to each step is quite large. 
Further each step of interfacing with different resources increases the opportunity for transcription errors.

@@

Here we provide a description of the BarnebyLives R package. 
BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert). 
Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert). 
Most importantly, Rupert was remembered as an individual entirely generous with there time to assist younger botanists with the more arcane aspects of field botany and taxonomy (@holmgren2017).

@@

## Usage
All steps of BarnebyLives except for label generation are run from within Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all of the on-disk memory associated with these data are for storing geo-spatial information, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is available in the package documentation. 
Functions which require the on-disk data require a path to the data as an argument. 
Manually supplying the argument allows for the users to judiciously decide a storage location suitable for there needs.  
We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
While these steps are run through bash, we have wrapped them in a R functions which bypass the need to enter the commands to a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites. 
Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s. 
The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`s) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).

@@

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives), and the stable version is available on CRAN. 
The package includes three real use-case vignettes (tutorials) on usage. 
One vignette "setting_up_files" explores setting up a instance for a certain geographic area. 
Another vignette "running_pipeline" showcases the usage of the package for processing data entered on a spreadsheet. 
A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels. 
All data used in this mansucript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript

>>>>

In fact innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative). 
The current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative, @davis2023herbarium). 
Further we anticipate that collections are yet to gain their most widespread utilization as a revitalization of natural history appears underway in ecology, fostered via novel approaches such as remote and electronic sensing, meta-barcoding, and community science  (@tosa2021rapid). 
While image or purely observational (rather than collections based) citizen science initializes (e.g. iNaturalist) have dovetailed to meet many needs of these studies specimens contain rich data which are not accessible via images. 
Namely specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes (@borges2020schrodinger), and seeds or pollen; eternally ensuing specimens as the ultimate data source to center most efforts around.

However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow (@prather2004decline). 
We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, than simply taking photographs via well-developed apps (@daru2018widespread, @mishler2020spatial). 
While many young botanists, capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria. 
Apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g. geology, geomorphology), nomenclature and Latin, various computer programs (e.g. Microsoft Office suite), and increasingly - plant systematics (@nanglu2023nature, @woodland2007botanists, @barrows2016crossroads). 
In the absence of suitable mentors this assuredly results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for many specimens to be accessioned at all, and increasingly ever collected.

@@

The generation of an herbarium specimen includes many steps which are easy to take for granted (@forman1989herbarium). 
For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate resources (printed topographic maps, or GIS software) at their disposal. 
In topographically complex areas, where borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their position. 
Finding appropriate sites names is another problem which can rarely be solved without a printed map, as many software maps now consider many features which would serve as site names extraneous in the era of GPS. 
The rate at which taxonomic innovations are occurring has left many Floras difficult for young users to interpret and has made it difficult for them to find more recently applied names.
Upon finding a name they may find it frustrating to hear that while published, the proposal has been accepted by few practitioners and they have unwittingly offended certain curators. 
Formatting a label correctly (e.g. abbreviations), if successful upon even setting up a mail merge, is a time consuming process and likely to introduce several errors in formatting. 
Even if a collector navigates all of these hurdles successfully, the time allocated to each step is quite large. 
Further each step of interfacing with different resources increases the opportunity for transcription errors.

@@

Here we provide a description of the BarnebyLives R package. 
BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing in 2000 (@welsh2001rupert). 
Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert). 
Most importantly, Rupert was remembered as being generous with his time to assist younger botanists with the more arcane aspects of field botany and taxonomy (@holmgren2017).

@@

## Usage
All steps of BarnebyLives except for label generation are run from within Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all of the on-disk memory associated with these data are for storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation. 
Functions which require the on-disk data require a path to the data as an argument.  
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.  
We anticipate most personal BarnebyLives instances will be less than several gigabytes (ours covering all of the conterminous Western U.S. is XX GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
While these steps are run through bash, we have wrapped them in R functions which bypass the need to enter the commands to a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites. 
Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s. 
The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`s) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).  
The total computer run time for processing these XXX specimens was XX minutes.

@@

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives), and the stable version is available on CRAN. 
The package includes three real use-case vignettes (tutorials) on usage. 
One vignette "setting_up_files" explores setting up a instance for a certain geographic area. 
Another vignette "running_pipeline" showcases the usage of the package for processing data entered on a spreadsheet. 
A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels. 
All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript
